---
title: "envs193ds final"
author: "Ella Menin"
format: pdf
---

# Setting up

```{r}
#| label: loading-in-packages
#| message: false
#| warning: false

# reading in package: tidyverse
library(tidyverse)
# reading in package: ggplot2
library(ggplot2)
# reading in package: rstatix
library(rstatix)
# reading in package: here
library(here)
# reading in package: readr
library(readr)
# reading in package: janitor
library(janitor)
```

```{r}
#| label: reading-in-nest-data

nests <- read_csv(here("data", "nest_data_final.csv"))

```


```{r}
#| label: reading-in-personal-data

my_data <- read_csv(here("data", "personal_data.csv"))

```


# Problem 1: Research writing

## a. Transparent statistical methods

* In part 1, my coworker used a linear model, as the response variable is the probability of American Avocet microhabitat use and the predictor variable is the distance to water edge. 

* In part 2, my coworker used a chi-square test of independence because they were testing for an association between two categorical variables: bird species and habitat type.

## b. Suggestions for rewriting

* Part 1: American Avocets were more likely to use some microhabitats than others depending on their distance from the water's edge. This shows that proximity to water is an important variable influencing where American Avocets occupy habitat.

* Part 2: The three bird species (American Avocet, Forster’s Tern, Black-necked Stilt) did not use habitat types in the same way. Habitat use patterns differed among species, showing that different shorebirds may prefer or rely on different habitats within the San Francisco Bay.

## c. Further analysis

One additional variable that could influence American Avocet microhabitat use is water depth (cm), as different birds prefer varying water depths depending on their characteristics. This could be measured at each survey location using a measurement tool, such as a meter stick, or by utilizing a water-depth gauge. In this specific context, water depth may affect habitat use because American Avocets feed by sweeping their bills through shallow water, so they may preferentially use areas with depths that maximize foraging efficiency.

# Problem 2: Data Analysis


# Problem 3: Affective and exploratory data

## a. Comparing visualizations

* My visualizations were very different from each other in the way that I represented my data. My exploratory visualizations focused more on showing visual patterns and trends between variables in my data. My affective visualization did that as well; however, it did it in a much more creative way, as it told a story about my entire day using my data as a tool to do so. There was much more story-telling involved in it. 

* All of my visualizations, both exploratory and my affective visualization, effectively demonstrated patterns regarding my water intake for the day and the conditions I experienced. My exploratory visualizations explicitly showed these through easy-to-read visualizations, while my affective visualization provided the viewer with multiple daily statistical summaries of my personal data that they could then compare. 

* Generally, I noticed in my exploratory visualizations that every variable I considered at least had some solid influence on my daily water intake. For my affective visualization, however, the main pattern shown is the fact that my water intake varies from day to day. So although these sound similar, one kind of visualization (exploratory) explicitly laid out the fact that the variables had relationships, the other kind of visualization (affective) showed the viewer the larger picture of the idea that my water intake is not a fixed value that remains the same every day. I believe this is because I created my affective visualization primarily with the intention of telling a story about my day and how heat and activity contributes to my water intake, rather than focusing on explicitly demonstrating how these variables vary day-by-day along with my water intake.

* I received excellent feedback from my peers! One of them suggested that I add in a couple more days of data into my visualization, which I did end up doing in the form of old posters laying on the floor. However, I still chose to highlight one main day of data, as I found it easier to tell a story about a single day by utilizing primarily that day's data. I basically added the other days so viewers could compare my water intakes depending on the day (I was inspired by calendar pages), but in the end, I chose to focus on a "day-in-my-life" narrative. 

## b. Designing an analysis

To determine whether daily water intake differed between days when I was sick and days when I was not sick, I am using a Wilcoxon rank-sum test. This analysis is appropriate because daily water intake (measured in L) is my response variable and sickness (sick or not sick) is a categorical predictor variable with two levels.

Side note: My personal prediction is that there is a correlation, and that I tended to drink more water on days I was sick than days I was not. 

## c. Checking assumptions and running an analysis

```{r}
#| label: cleaning-and-wrangling-data

# creating new objected for cleaned data using my_data data frame
my_data_clean <- my_data |> 
  # cleaning names: all lowercase and adding underscores
  clean_names() |>
  # removing an empty row i accidentally included
  remove_empty("rows") |>
  # converts sick status into categorical variable
  mutate(
    sick = factor(sick_y_n)
  )

```

```{r}
#| label: checking-assumptions

# looking at normal distribution using histogram
# base layer ggplot with cleaned data frame
ggplot(data = my_data_clean,
       # setting x axis: water intake in liters
       mapping = aes(x = water_intake_l)) +
  # first layer geom_histogram with 10 bins
  geom_histogram(bins = 10) +
  # relabeling x and y axis
  labs(x = "Daily Water Intake (L)", # x axis is daily water intake in liters
       y = "Frequency") # y is frequency (number observations in each bin)

# looking at distribution using a qq plot
# base layer ggplot with cleaned data frame
ggplot(data = my_data_clean,
       # looking at water intake in liters
       mapping = aes(sample = water_intake_l)) +
  # adding individual observations
  stat_qq() +
  # adding in reference line to assess normality of observations
  stat_qq_line()

```

I examined the distribution of daily water intake using a histogram and a Q-Q plot. The histogram indicated that water intake was left-skewed, and the Q-Q plot showed deviations from the reference line in the tails, suggesting that the data were not normally distributed. Because the assumption of normality was not met, a Wilcoxon rank-sum test is still appropriate to compare water intake between sick and non-sick days.

```{r}
#| label: performing-wilcox-test

# creating new object from running the wilcoxon rank-sum test
wilcox_results <- wilcox.test(
  # examining correlation between water intake and sickness
  water_intake_l ~ sick,
  # using my cleaned data data frame
  data = my_data_clean
)

# displaying results
wilcox_results

```

```{r}
#| label: calculating-effect-size

# calculating effect size
effect_size <- wilcox_effsize(
  # my cleaned data frame
  my_data_clean,
  # examining correlation between water intake and sickness
  water_intake_l ~ sick
)

# displaying results
effect_size

```

## d. Create a visualization

```{r}
#| label: creating-boxplot

# base layer ggplot with cleaned data frame
ggplot(data = my_data_clean,
       # setting x and y axes
       mapping = aes(x = sick, # x axis is sickness
           y = water_intake_l, # y axis is water intake in liters
           # fill color by sickness yes or no
           fill = sick)) +
  # first layer boxplot
  geom_boxplot(alpha = 0.7, # alpha 0.7 so you can still see individual observations through boxplot
               # taking out extreme outliers that may mess things up
               outlier.shape = NA) +
  # second layer jitterplot to show individual observations
  geom_jitter(width = 0.15, # controlling width
              size = 2, # size 2 for easy readability
              height = 0, # controlling height
              alpha = 0.7) + # alpha 0.7 so you can see boxplot too
  # setting custom colors
  scale_fill_manual(values = c("deepskyblue3", # not sick = pretty blue
                               "tomato1")) + # sick = red
  # setting axes titles
  labs(
    # x axis is sickness status (sick or not sick)
    x = "Sickness Status",
    # y is daily water intake measured in liters
    y = "Daily Water Intake (L)"
  ) +
  # choosing a nice, simple theme
  theme_bw()

```

## e. Figure caption

Figure 1. Daily water intake on days when I was sick and days when the I was not sick. Boxplots show the median and interquartile range of daily water intake (measured in liters) for each sickness category (yes and no). Individual observations are displayed as black jittered points to show the distribution of measurements. The data around the blue box indicates days when the participant was not sick and those around light red box indicates days when the participant was sick.

## f. Write about your results

Daily water intake differed significantly between days when I was sick and days when I was not sick (Wilcoxon rank-sum test, W = 50.5, p = 0.025). As shown in Figure 1, the median water intake was higher on sick days than on non-sick days, supporting my prediction that I would drink more water when ill. Sickness status was associated with a moderate effect on water consumption (r = 0.362), suggesting that the difference was not only statistically significant but also meaningful in magnitude.