---
title: "envs193ds final"
author: "Ella Menin"
format: pdf
---

# Setting up

```{r}
#| label: loading-in-packages
#| message: false
#| warning: false

# reading in package: tidyverse
library(tidyverse)
# reading in package: ggplot2
library(ggplot2)
# reading in package: rstatix
library(rstatix)
# reading in package: here
library(here)
# reading in package: readr
library(readr)
# reading in package: janitor
library(janitor)
```

```{r}
#| label: reading-in-nest-data

```


```{r}
#| label: reading-in-personal-data

my_data <- read_csv(here("data", "personal_data.csv"))

```


# Problem 1: Research writing

## a. Transparent statistical methods

* In part 1, my coworker used a linear model, as the response variable is the probability of American Avocet microhabitat use and the predictor variable is the distance to water edge. 

* In part 2, my coworker used a chi-square test of independence because they were testing for an association between two categorical variables: bird species and habitat type.

## b. Suggestions for rewriting

* Part 1: American Avocets were more likely to use some microhabitats than others depending on their distance from the water's edge. This shows that proximity to water is an important variable influencing where American Avocets occupy habitat.

* Part 2: The three bird species (American Avocet, Forster’s Tern, Black-necked Stilt) did not use habitat types in the same way. Habitat use patterns differed among species, showing that different shorebirds may prefer or rely on different habitats within the San Francisco Bay.

## c. Further analysis

One additional variable that could influence American Avocet microhabitat use is water depth (cm), as different birds prefer varying water depths depending on their characteristics. This could be measured at each survey location using a measurement tool, such as a meter stick, or by utilizing a water-depth gauge. In this specific context, water depth may affect habitat use because American Avocets feed by sweeping their bills through shallow water, so they may preferentially use areas with depths that maximize foraging efficiency.

# Problem 2: Data Analysis


# Problem 3: Affective and exploratory data

## a. Comparing visualizations

* My visualizations were very different from each other in the way that I represented my data. My exploratory visualizations focused more on showing visual patterns and trends between variables in my data. My affective visualization did that as well; however, it did it in a much more creative way, as it told a story about my entire day using my data as a tool to do so. There was much more story-telling involved in it. 

* All of my visualizations, both exploratory and my affective visualization, effectively demonstrated patterns regarding my water intake for the day and the conditions I experienced. My exploratory visualizations explicitly showed these through easy-to-read visualizations, while my affective visualization provided the viewer with multiple daily statistical summaries of my personal data that they could then compare. 

* Generally, I noticed in my exploratory visualizations that every variable I considered at least had some solid influence on my daily water intake. For my affective visualization, however, the main pattern shown is the fact that my water intake varies from day to day. So although these sound similar, one kind of visualization (exploratory) explicitly laid out the fact that the variables had relationships, the other kind of visualization (affective) showed the viewer the larger picture of the idea that my water intake is not a fixed value that remains the same every day. I believe this is because I created my affective visualization primarily with the intention of telling a story about my day and how heat and activity contributes to my water intake, rather than focusing on explicitly demonstrating how these variables vary day-by-day along with my water intake.

* I received excellent feedback from my peers! One of them suggested that I add in a couple more days of data into my visualization, which I did end up doing in the form of old posters laying on the floor. However, I still chose to highlight one main day of data, as I found it easier to tell a story about a single day by utilizing primarily that day's data. I basically added the other days so viewers could compare my water intakes depending on the day (I was inspired by calendar pages), but in the end, I chose to focus on a "day-in-my-life" narrative. 

## b. Designing an analysis

To determine whether daily water intake differed between days when I was sick and days when I was not sick, I am using a Wilcoxon rank-sum test. This analysis is appropriate because daily water intake (measured in L) is my response variable and sickness (sick or not sick) is a categorical predictor variable with two levels.

## c. Checking assumptions and running an analysis

```{r}
#| label: cleaning-and-wrangling-data

# creating new objected for cleaned data using my_data data frame
my_data_clean <- my_data |> 
  # cleaning names: all lowercase and adding underscores
  clean_names() |>
  # removing an empty row i accidentally included
  remove_empty("rows") |>
  # converts sick status into categorical variable
  mutate(
    sick = factor(sick_y_n)
  )

```

